---
title: "trace gas oxidation genes in metagenome assembly"
author: "Vojtech Tlaskal"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    latex-engine: xelatex      
    geometry: margin=1in
---

# Data Processing Pipeline

This report combines R and bash commands to process and analyze assembled and gene-called metagenomic data from subarctic grassland soil in order to identify target trace gas oxidation genes and their taxonomy.

```{r setup}
# Load required libraries
library(tidyverse)
library(tidylog)
library(readr)
library(readxl)
library(openxlsx2)

# Set global options
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Raw FASTQ QC
```{bash}
# Trimmomatic
for Sample in *_R1.fastq
do
    base=$(basename ${Sample} _R1.fastq)
    trimmomatic PE -threads 20 -trimlog trimLogFile -summary statsSummaryFile \
    ${Sample} ${base}_R2.fastq \
    ${base}_R1.trim.fastq ${base}_R1un.trim.fastq \
    ${base}_R2.trim.fastq ${base}_R2un.trim.fastq \
    ILLUMINACLIP:TruSeq3-PE-2.fa:2:30:10:2:True LEADING:3 TRAILING:3 MINLEN:36
done

# Fastp, loop over all pairs of input fastq files in the directory
for R1_file in ${fastq_dir}/*_R1.fastq
do
    R2_file=${R1_file%_R1.fastq}_R2.fastq
    base=$(basename ${R1_file} _R1.fastq)

    # Run fastp on the pair of files
    fastp --in1 ${R1_file} \
          --in2 ${R2_file} \
          --out1 ${output_dir}/${base}.trim_R1.fastq \
          --out2 ${output_dir}/${base}.trim_R2.fastq \
          --qualified_quality_phred 4 \
          --length_required 31 \
          --correction \
          --json ${output_dir}/${base}.trim.json \
          --html ${output_dir}/${base}.trim.html
done
```

## Metagenome assembly, genecalling and annotation 

```{bash}
# MEGAHIT assembly
# Run assembly commands for each sample
cd A1 && megahit -1 Ambient1_S1_L001_R1.trim.fastq -2 Ambient1_S1_L001_R2.trim.fastq -o ASSEMBLY -t 12 && cd ..
cd A2 && megahit -1 Ambient2_S2_L001_R1.trim.fastq -2 Ambient2_S2_L001_R2.trim.fastq -o ASSEMBLY -t 12 && cd ..
cd A4 && megahit -1 Ambient4_S3_L001_R1.trim.fastq -2 Ambient4_S3_L001_R2.trim.fastq -o ASSEMBLY -t 12 && cd ..
cd A5 && megahit -1 Ambient5_S4_L001_R1.trim.fastq -2 Ambient5_S4_L001_R2.trim.fastq -o ASSEMBLY -t 12 && cd ..
cd W1 && megahit -1 Warm1_S6_L001_R1.trim.fastq -2 Warm1_S6_L001_R2.trim.fastq -o ASSEMBLY -t 12 && cd ..
cd W2 && megahit -1 Warm2_S7_L001_R1.trim.fastq -2 Warm2_S7_L001_R2.trim.fastq -o ASSEMBLY -t 12 && cd ..
cd W4 && megahit -1 Warm4_S8_L001_R1.trim.fastq -2 Warm4_S8_L001_R2.trim.fastq -o ASSEMBLY -t 12 && cd ..
cd W5 && megahit -1 Warm5_S9_L001_R1.trim.fastq -2 Warm5_S9_L001_R2.trim.fastq -o ASSEMBLY -t 12

# Gene prediction using Prodigal
# Define an array of sample names
samples=("A1" "A2" "A4" "A5" "W1" "W2" "W4" "W5")
mkdir -p All_Iceland_proteins

# Loop through each sample and run Prodigal
for sample in "${samples[@]}"
do
    # Set input file path and output directory
    input_file="${sample}/ASSEMBLY/final.contigs.fa"
    output_dir="All_Iceland_proteins"
    output_file="${sample}.proteins.faa"

    # Run Prodigal for gene prediction
    prodigal -i "${input_file}" -o "${output_dir}/${sample}" -a "${output_dir}/${output_file}" -p meta
done

# Diamond annotation
# Download necessary files for taxonomic mapping
# --taxonmap: File for mapping accession numbers to taxids
# --taxonnodes, --taxonnames: Taxonomic nodes and names
# Funcgenes_51_Sep2023 from Greening lab metabolic marker gene databases via 
# Leung, Pok Man; Greening, Chris (2020). Greening lab metabolic marker gene databases. Monash University. Collection. https://doi.org/10.26180/c.5230745

# Process all downloaded gzipped files to create DIAMOND database
for file in *.gz; do
    diamond makedb --in $file --db ../Diamond_database/Funcgenes_51_Sep2023 \
    --taxonmap ../prot.accession2taxid.FULL.gz \
    --taxonnodes ../taxdmp/nodes.dmp \
    --taxonnames ../taxdmp/names.dmp \
    --threads 5
done

# Create output directory for annotations
mkdir -p MAG_Contig_Annotation
cd All_Iceland_proteins

# Set DIAMOND database path
database_path="Funcgenes_51_Sep2023.dmnd"

# Loop through all .faa files and run DIAMOND blastp
for input_file in *.faa; do
    # Define output file path
    output_file="../MAG_Contig_Annotation/${input_file%.faa}_Annotation_file.txt"
    
    # Run DIAMOND blastp
    diamond blastp -p 10 -d "$database_path" -q "$input_file" -o "$output_file" \
    -k 1 --outfmt 6 qseqid sseqid pident length mismatch gapopen qstart qend \
    sstart send evalue bitscore qcovhsp salltitles staxids
done

```

## Data Preparation

```{r data_processing}
# Extract gene-called sequences
# Define a function to read, add columns, and process data
read_add_and_process_data <- function(filename) {
  data <- read_tsv(filename, 
                   col_names = c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", 
                                 "qend", "sstart", "send", "evalue", "bitscore", 
                                 "qcovhsp", "salltitles"))
  
  # Add Gene and AccessionNumber columns by splitting sseqid
  data <- data %>%
    separate(sseqid, into = c("Gene", "AccessionNumber"), sep = "-", remove = FALSE) %>%
    filter(!grepl("^[A-Za-z]+$", AccessionNumber))
  
  data <- data %>%
    filter(pident > 75 & qcovhsp > 70) %>%
    distinct(qseqid, .keep_all = TRUE) %>%
    glimpse()
  
  return(data)
}

# Read, add columns, and process data for each dataset
A1 <- read_add_and_process_data("A1.proteins_Annotation_file.txt.gz")
A2 <- read_add_and_process_data("A2.proteins_Annotation_file.txt.gz")
A4 <- read_add_and_process_data("A4.proteins_Annotation_file.txt.gz")
A5 <- read_add_and_process_data("A5.proteins_Annotation_file.txt.gz")
W1 <- read_add_and_process_data("W1.proteins_Annotation_file.txt.gz")
W2 <- read_add_and_process_data("W2.proteins_Annotation_file.txt.gz")
W4 <- read_add_and_process_data("W4.proteins_Annotation_file.txt.gz")
W5 <- read_add_and_process_data("W5.proteins_Annotation_file.txt.gz")

# Add a sample column 
A1$Sample <- "Ambient1"
A2$Sample <- "Ambient2"
A4$Sample <- "Ambient4"
A5$Sample <- "Ambient5"
W1$Sample <- "Warm1"
W2$Sample <- "Warm2"
W4$Sample <- "Warm4"
W5$Sample <- "Warm5"

# Combine the data frames into a single data frame
combined_data <- rbind(A1, A2, A4, A5, W1, W2, W4, W5)

# load target annotations
diamond_annotations <- read_csv("diamond_target_annotations.csv", 
    trim_ws = FALSE) %>% 
  janitor::clean_names() %>% 
  glimpse()

test_duplicates <- distinct(combined_data, qseqid)

# combine with targets
genes_list <- combined_data %>% 
  janitor::clean_names() %>% 
  inner_join(diamond_annotations, "gene") %>% 
  glimpse

distinct <- distinct(genes_list, qseqid)
distinct_full <- distinct(genes_list, qseqid, .keep_all = T) |> 
  glimpse()

#write_xlsx(distinct, "gene_list.xlsx")
#write_xlsx(distinct_full, "genes_list_full.xlsx")

# Get list of genes in fasta
grouped_data <- distinct_full %>% 
  select(sample, qseqid) |> 
  group_by(sample)

# Split the data into a list of tibbles
grouped_tibbles <- group_split(grouped_data)

# Iterate over the list and write each tibble to a separate text file
for (i in seq_along(grouped_tibbles)) {
  # Extract the group name from the first row of the tibble
  group_name <- grouped_tibbles[[i]]$sample[1]
  
  # Write the tibble to a text file, including the group name in the filename
  write.table(grouped_tibbles[[i]], paste0("group_", group_name, ".txt"), sep = "\t", row.names = FALSE, quote = F)
}
```

## Fasta file management

Bash commands to filter fasta files:

```{bash}
# filter target genes for CO/CO2/H2
for file in group*.txt
do
sample=${file%%.txt}
awk 'NR>1 {print $2}' ${file} > ${sample}_edit.txt
done

# seqkit to filter sequences from fasta gene-calls based on list of target genes
seqkit grep -n -f group_Ambient1_edit.txt A1_trimmed.fas > A1_trace_genes.fas 
seqkit grep -n -f group_Ambient2_edit.txt A2_trimmed.fas > A2_trace_genes.fas
seqkit grep -n -f group_Ambient4_edit.txt A4_trimmed.fas > A4_trace_genes.fas
seqkit grep -n -f group_Ambient5_edit.txt A5_trimmed.fas > A5_trace_genes.fas

seqkit grep -n -f group_Warm1_edit.txt W1_trimmed.fas > W1_trace_genes.fas
seqkit grep -n -f group_Warm2_edit.txt W2_trimmed.fas > W2_trace_genes.fas
seqkit grep -n -f group_Warm4_edit.txt W4_trimmed.fas > W4_trace_genes.fas
seqkit grep -n -f group_Warm5_edit.txt W5_trimmed.fas > W5_trace_genes.fas

# append name
for file in *_trace_genes.fas
do
   sample=${file%%_trace_genes.fas}
awk -v fname="$sample" '/^>/ {$0=$0 "|" fname}1' ${file} > ${sample}_trace_genes_renamed.fas
done

# merge all trace gas oxidation genes
cat *_trace_genes_renamed.fas > all_trace_genes_renamed.fas
# translate trace gas oxidation genes
seqkit translate all_trace_genes_renamed.fas --trim -T 11 > all_trace_genes_renamed.faa
```

## Download and process RefSeq bacterial genomes

```{bash}
# RefSeq bacteria download using datasets package
datasets download genome taxon bacteria --assembly-source RefSeq --dehydrated --include protein
unzip ncbi_dataset.zip -d refseq_bacteria
datasets rehydrate --directory refseq_bacteria

# rename files
for file in */*.faa; do
 mv -- "$file" "${file%/*.faa}/${file%/*.faa}"
done

# copy all to single directory
mkdir all
for file in */*; do
   cp -n "$file" all/
done

# insert filename to sequence header separated by |
for file in *.1; do
 awk -v file="${file%.*}" 'BEGIN{FS=OFS="|"} /^>/{$0=">"file"|"substr($0,2)} 1' "$file" > "${file%.*}"_modified.faa
done
for file in *.2; do
 awk -v file="${file%.*}" 'BEGIN{FS=OFS="|"} /^>/{$0=">"file"|"substr($0,2)} 1' "$file" > "${file%.*}"_modified.faa
done
for file in *.3; do
 awk -v file="${file%.*}" 'BEGIN{FS=OFS="|"} /^>/{$0=">"file"|"substr($0,2)} 1' "$file" > "${file%.*}"_modified.faa
done
for file in *.4; do
 awk -v file="${file%.*}" 'BEGIN{FS=OFS="|"} /^>/{$0=">"file"|"substr($0,2)} 1' "$file" > "${file%.*}"_modified.faa
done
for file in *.5; do
 awk -v file="${file%.*}" 'BEGIN{FS=OFS="|"} /^>/{$0=">"file"|"substr($0,2)} 1' "$file" > "${file%.*}"_modified.faa
done
# cat for this number of args
find . -maxdepth 1 -type f -name '*_modified.faa' -print0 |
  sort -z |
  xargs -0 cat -- >> merged_bacteria.faa
# gives 324508 accessions, all merged to one file
```

## Diamond against RefSeq to infer taxonomy of target genes

```{bash}
# creating a diamond-formatted database file
diamond makedb --in merged_bacteria.faa -d merged_bacteria
# running a search in blastp mode
diamond blastp -d merged_bacteria.dmnd -q all_trace_genes_renamed.faa --threads 50 --outfmt 6 -v -o all_trace_genes_renamed_diamond.tsv
```

## Process taxonomy

```{r taxonomy}
# prepare taxonomy
refseq_tax <- read_delim("ar53_taxonomy_r214.tsv", # official RefSeq archaeal taxonomy
                         delim = "\t", escape_double = FALSE, 
                         col_names = FALSE) %>%
  add_row(read_delim("bac120_taxonomy_r214.tsv", # official RefSeq bacterial taxonomy
                     delim = "\t", escape_double = FALSE, 
                     col_names = FALSE)) %>% 
  mutate(accession_helper = str_remove(X1, "RS_")) %>% 
  select(-X1) %>% 
  mutate(accession_helper = str_remove(accession_helper, "GB_")) |> 
  rename(accession = accession_helper) |> 
  tidyr::separate_wider_delim(accession, delim = ".", names = c("accession_helper", "accession_version")) |>
  distinct(accession_helper, .keep_all = T)
# load RefSeq tax, source is custom GTDB-Tk analysis
gtdb_tk_tax <- read_csv("gtdbtk.bac120.summary.csv", 
    trim_ws = FALSE) |>  
  select(user_genome, classification) |> 
  dplyr::rename(X2 = classification) |> 
  tidyr::separate_wider_delim(user_genome, delim = ".", names = c("accession_helper", "accession_version")) |>
  distinct(accession_helper, .keep_all = T)
tax_merged <- add_row(refseq_tax, gtdb_tk_tax)

# diamond results
all_trace_genes_renamed_diamond <- read_delim("all_trace_genes_renamed_diamond.tsv", 
                                              delim = "\t", escape_double = FALSE, 
                                              col_names = FALSE) |> 
  group_by(X1) |> 
  slice_max(X12, with_ties = T) |> # bitscore filter
  slice_min(X11, with_ties = T) |> # evalue filter
  slice_max(X3, with_ties = F, n = 1) |>  # pident filter, ties F to choose the first one if previous are same
  ungroup()

# merge gene list and diamond results, filter for trace gases
distinct_full_taxonomy <- select(all_trace_genes_renamed_diamond, X1, X2, X11, X12) |> 
  tidyr::separate_wider_delim(X1, delim = "|", names = c("qseqid", "treatment")) |> 
  tidyr::separate_wider_delim(X2, delim = "|", names = c("accession_helper", "gene_id")) |>
  left_join(tax_merged, by = "accession_helper") |> 
  right_join(distinct_full, "qseqid") |> 
  tidyr::separate_wider_delim(X2, delim = ";", names = c("domain", "phylum", "class","order", "family", "genus", "species")) |>
  filter(function_diamond == "Carbon monoxide oxidation" | 
           function_diamond == "Methane oxidation" | 
           function_diamond == "hydrogenase") |>
  group_by(gene, class, treatment) |>  
  summarise(n = n()) |> 
  #filter(is.na(class)) |> # see how many does not have taxonomy
  left_join(read_csv("gene_numbers.csv", 
    trim_ws = FALSE), # gene-call counts
            by = "treatment") |>
  mutate(perc = (n/gene_count)*1e6) |> # genes per mio of genes
  filter(!is.na(class)) |> # remove no hits from diamond tax
  left_join(read_csv("distinct_full_taxonomy_helper.csv", 
    trim_ws = FALSE),
            by = c("gene", "class")) |># join helper table to plot only most abundant
  mutate(class_helper = if_else(is.na(class_helper), true = "others", false = class_helper)) |>
  group_by(gene, class_helper, temp) |> 
  summarise(perc_mean = mean(perc), se = sd(perc)/(sqrt(n())), n = n()) |>
  ungroup() |> 
  group_by(class_helper) |> 
  mutate(sort_help = mean(perc_mean)) |> # display in ggplot
  mutate(class_helper = str_remove(class_helper, "c__")) |> 
  mutate(temp = fct_recode(temp, `Long−term warmed` = "warm")) |>
  mutate(temp = fct_recode(temp, Ambient = "ambient")) |>
  mutate(wd = if_else(gene == "CoxL", true = 1, false = 
                        if_else(gene == "NiFe", true = 1, false = 0.3))) |>
  glimpse()
```


