---
title: "trace gas oxidation genes in public metagenome assembly"
author: "Vojtech Tlaskal"
date: today
format:
  pdf:
    toc: true
    number-sections: true
    latex-engine: xelatex      
    geometry: margin=1in
---

# Data Processing Pipeline

This report combines R and bash commands to process and analyze assembled and gene-called metagenomic data from subarctic grassland soil in order to identify target trace gas oxidation genes and their taxonomy. Two public metagenomes from the NCBI BioProject PRJNA746424 with BioSample SAMN20206491 and SAMN20206488 are analysed.

```{r setup}
# Load required libraries
library(tidyverse)
library(tidylog)
library(readr)
library(readxl)
library(openxlsx2)

# Set global options
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Raw FASTQ QC
```{bash}
# download using fasterq dump
for acc in SRR15130125 SRR15130128; do sratoolkit.3.0.0-ubuntu64/bin/fasterq-dump --split-files $acc; done

# Trimmomatic
for Sample in *_R1.fastq
do
    base=$(basename ${Sample} _R1.fastq)
    trimmomatic PE -threads 20 -trimlog trimLogFile -summary statsSummaryFile \
    ${Sample} ${base}_R2.fastq \
    ${base}_R1.trim.fastq ${base}_R1un.trim.fastq \
    ${base}_R2.trim.fastq ${base}_R2un.trim.fastq \
    ILLUMINACLIP:TruSeq3-PE-2.fa:2:30:10:2:True LEADING:3 TRAILING:3 MINLEN:36
done

# Fastp, loop over all pairs of input fastq files in the directory
for R1_file in ${fastq_dir}/*_R1.fastq
do
    R2_file=${R1_file%_R1.fastq}_R2.fastq
    base=$(basename ${R1_file} _R1.fastq)

    # Run fastp on the pair of files
    fastp --in1 ${R1_file} \
          --in2 ${R2_file} \
          --out1 ${output_dir}/${base}.trim_R1.fastq \
          --out2 ${output_dir}/${base}.trim_R2.fastq \
          --qualified_quality_phred 4 \
          --length_required 31 \
          --correction \
          --json ${output_dir}/${base}.trim.json \
          --html ${output_dir}/${base}.trim.html
done
```

## Metagenome assembly, genecalling and annotation 

```{bash}
# MEGAHIT assembly
# Run assembly commands for each sample
cd A6 && megahit -1 SRR15130125_R1.trim.fastq -2 SRR15130125_R2.trim.fastq -o ASSEMBLY -t 12 && cd ..
cd W6 && megahit -1 SRR15130128_R1.trim.fastq -2 SRR15130128_R2.trim.fastq -o ASSEMBLY -t 12

# Gene prediction using Prodigal
# Define an array of sample names
samples=("A6" "W6")
mkdir -p All_Iceland_proteins

# Loop through each sample and run Prodigal
for sample in "${samples[@]}"
do
    # Set input file path and output directory
    input_file="${sample}/ASSEMBLY/final.contigs.fa"
    output_dir="All_Iceland_proteins"
    output_file="${sample}.proteins.faa"

    # Run Prodigal for gene prediction
    prodigal -i "${input_file}" -o "${output_dir}/${sample}" -a "${output_dir}/${output_file}" -p meta
done

# Diamond annotation
# Download necessary files for taxonomic mapping
# --taxonmap: File for mapping accession numbers to taxids
# --taxonnodes, --taxonnames: Taxonomic nodes and names
# Funcgenes_51_Sep2023 from Greening lab metabolic marker gene databases via 
# Leung, Pok Man; Greening, Chris (2020). Greening lab metabolic marker gene databases. Monash University. Collection. https://doi.org/10.26180/c.5230745

# Process all downloaded gzipped files to create DIAMOND database
for file in *.gz; do
    diamond makedb --in $file --db ../Diamond_database/Funcgenes_51_Sep2023 \
    --taxonmap ../prot.accession2taxid.FULL.gz \
    --taxonnodes ../taxdmp/nodes.dmp \
    --taxonnames ../taxdmp/names.dmp \
    --threads 5
done

database_path="./Database/Funcgenes_51_Dec2021.dmnd"

# Loop through all .faa files and run DIAMOND blastp
for input_file in *.faa; do
    # Define output file path
    output_file="./gene_annotation/${input_file%.faa}_Annotation_file.txt"
    
    # Run DIAMOND blastp
    diamond blastp -p 30 -d "$database_path" -q "$input_file" -o "$output_file" \
    -k 1 --outfmt 6 qseqid sseqid pident length mismatch gapopen qstart qend \
    sstart send evalue bitscore qcovhsp salltitles staxids
done
```

## Data Preparation

```{r data_processing}
# Extract gene-called sequences
# Define a function to read, add columns, and process data
read_add_and_process_data <- function(filename) {
  data <- read_tsv(filename, 
                   col_names = c("qseqid", "sseqid", "pident", "length", "mismatch", "gapopen", "qstart", 
                                 "qend", "sstart", "send", "evalue", "bitscore", 
                                 "qcovhsp", "salltitles"))
  
  # Add Gene and AccessionNumber columns by splitting sseqid
  data <- data %>%
    separate(sseqid, into = c("Gene", "AccessionNumber"), sep = "-", remove = FALSE) %>%
    filter(!grepl("^[A-Za-z]+$", AccessionNumber))
  
  data <- data %>%
    filter(pident > 75 & qcovhsp > 70) %>%
    distinct(qseqid, .keep_all = TRUE) %>%
    glimpse()
  
  return(data)
}

# Read, add columns, and process data for each dataset
A6 <- read_add_and_process_data("A6.proteins_Annotation_file.txt.gz")
W6 <- read_add_and_process_data("W6.proteins_Annotation_file.txt.gz")

# Add a sample column 
A6$Sample <- "Ambient6"
W6$Sample <- "Warm6"

# Combine the data frames into a single data frame
combined_data_mg_2016 <- rbind(A6, W6)

# load target annotations
diamond_annotations <- read_csv("diamond_target_annotations.csv", 
    trim_ws = FALSE) %>% 
  janitor::clean_names() %>% 
  glimpse()

test_duplicates <- distinct(combined_data_mg_2016, qseqid)

# combine with targets
genes_list_2016 <- combined_data_mg_2016 %>% 
  janitor::clean_names() %>% 
  inner_join(diamond_annotations, "gene") %>% 
  glimpse

distinct_2016 <- distinct(genes_list_2016, qseqid)
distinct_full_2016 <- distinct(genes_list_2016, qseqid, .keep_all = T) %>% 
  glimpse()

# Get list of genes in fasta
grouped_data <- distinct_full %>% 
  select(sample, qseqid) |> 
  group_by(sample)

# Split the data into a list of tibbles
grouped_tibbles <- group_split(grouped_data)

# Iterate over the list and write each tibble to a separate text file
for (i in seq_along(grouped_tibbles)) {
  # Extract the group name from the first row of the tibble
  group_name <- grouped_tibbles[[i]]$sample[1]
  
  # Write the tibble to a text file, including the group name in the filename
  write.table(grouped_tibbles[[i]], paste0("group_", group_name, ".txt"), sep = "\t", row.names = FALSE, quote = F)
}
```

## Fasta file management

Bash commands to filter fasta files:

```{bash}
# filter target genes for CO/CO2/H2
for file in group*.txt
do
sample=${file%%.txt}
awk 'NR>1 {print $2}' ${file} > ${sample}_edit.txt
done

# trim fasta header
for file in *_final.contigs.proteins.faa
do
sample=${file%%_final.contigs.proteins.faa}
cut -d" " -f1 ${file} > ${sample}_trimmed.fas
done

# seqkit to filter sequences from fasta gene-calls based on list of target genes
seqkit grep -n -f group_Ambient6_edit.txt SRR15130125_trimmed.fas > A6_trace_genes.faa &&
seqkit grep -n -f group_Warm6_edit.txt SRR15130128_trimmed.fas > W6_trace_genes.faa

# append name
for file in *_trace_genes.fas
do
   sample=${file%%_trace_genes.fas}
awk -v fname="$sample" '/^>/ {$0=$0 "|" fname}1' ${file} > ${sample}_trace_genes_renamed.fas
done

# merge all trace gas oxidation genes
cat *_trace_genes_renamed.fas > all_trace_genes_renamed.fas
# translate trace gas oxidation genes
seqkit translate all_trace_genes_renamed.fas --trim -T 11 > all_trace_genes_renamed.faa
```

## Diamond against RefSeq to infer taxonomy of target genes

```{bash}
# creating a diamond-formatted database file
diamond makedb --in merged_bacteria.faa -d merged_bacteria
# running a search in blastp mode
diamond blastp -d merged_bacteria.dmnd -q all_trace_genes_renamed_2016.faa --threads 50 --outfmt 6 -v -o all_trace_genes_renamed_diamond_2016.tsv
```

## Process taxonomy

```{r taxonomy}
# prepare taxonomy
refseq_tax <- read_delim("ar53_taxonomy_r214.tsv", # official RefSeq archaeal taxonomy
                         delim = "\t", escape_double = FALSE, 
                         col_names = FALSE) %>%
  add_row(read_delim("bac120_taxonomy_r214.tsv", # official RefSeq bacterial taxonomy
                     delim = "\t", escape_double = FALSE, 
                     col_names = FALSE)) %>% 
  mutate(accession_helper = str_remove(X1, "RS_")) %>% 
  select(-X1) %>% 
  mutate(accession_helper = str_remove(accession_helper, "GB_")) |> 
  rename(accession = accession_helper) |> 
  tidyr::separate_wider_delim(accession, delim = ".", names = c("accession_helper", "accession_version")) |>
  distinct(accession_helper, .keep_all = T)
# load RefSeq tax, source is custom GTDB-Tk analysis
gtdb_tk_tax <- read_csv("gtdbtk.bac120.summary.csv", 
    trim_ws = FALSE) |>  
  select(user_genome, classification) |> 
  dplyr::rename(X2 = classification) |> 
  tidyr::separate_wider_delim(user_genome, delim = ".", names = c("accession_helper", "accession_version")) |>
  distinct(accession_helper, .keep_all = T)
tax_merged <- add_row(refseq_tax, gtdb_tk_tax)

# diamond results 
all_trace_genes_renamed_diamond_2016 <- read_delim("all_trace_genes_renamed_diamond_2016.tsv.gz", 
                                              delim = "\t", escape_double = FALSE, 
                                              col_names = FALSE) %>% 
  group_by(X1) %>% 
  slice_max(X12, with_ties = T) %>% # bitscore filter
  slice_min(X11, with_ties = T) %>% # evalue filter
  slice_max(X3, with_ties = F, n = 1) %>%  # pident filter, ties F to choose the first one if previous are same
  ungroup()

# merge gene list and diamond results, filter for trace gases
distinct_full_taxonomy <- select(all_trace_genes_renamed_diamond_2016, X1, X2, X11, X12) |> 
  tidyr::separate_wider_delim(X1, delim = "|", names = c("qseqid", "treatment")) |> 
  tidyr::separate_wider_delim(X2, delim = "|", names = c("accession_helper", "gene_id")) |>
  left_join(tax_merged, by = "accession_helper") |> 
  right_join(distinct_full, "qseqid") |> 
  tidyr::separate_wider_delim(X2, delim = ";", names = c("domain", "phylum", "class","order", "family", "genus", "species")) |>
  filter(function_diamond == "Carbon monoxide oxidation" | 
           function_diamond == "Methane oxidation" | 
           function_diamond == "hydrogenase") |>
  group_by(gene, class, treatment) |>  
  summarise(n = n()) |> 
  #filter(is.na(class)) |> # see how many does not have taxonomy
  left_join(read_csv("gene_numbers.csv", 
    trim_ws = FALSE), # gene-call counts
            by = "treatment") |>
  mutate(perc = (n/gene_count)*1e6) |> # genes per mio of genes
  filter(!is.na(class)) |> # remove no hits from diamond tax
  left_join(read_csv("distinct_full_taxonomy_helper.csv", 
    trim_ws = FALSE),
            by = c("gene", "class")) |># join helper table to plot only most abundant
  mutate(class_helper = if_else(is.na(class_helper), true = "others", false = class_helper)) |>
  group_by(gene, class_helper, temp) |> 
  summarise(perc_mean = mean(perc), se = sd(perc)/(sqrt(n())), n = n()) |>
  ungroup() |> 
  group_by(class_helper) |> 
  mutate(sort_help = mean(perc_mean)) |> # display in ggplot
  mutate(class_helper = str_remove(class_helper, "c__")) |> 
  mutate(temp = fct_recode(temp, `Long−term warmed` = "warm")) |>
  mutate(temp = fct_recode(temp, Ambient = "ambient")) |>
  mutate(wd = if_else(gene == "CoxL", true = 1, false = 
                        if_else(gene == "NiFe", true = 1, false = 0.3))) |>
  glimpse()

# get individual data for point plotting together with MGs from 2022
individual_data_2016 <- select(all_trace_genes_renamed_diamond_2016, X1, X2, X11, X12) %>% 
  tidyr::separate_wider_delim(X1, delim = "|", names = c("qseqid", "treatment")) %>% 
  tidyr::separate_wider_delim(X2, delim = "|", names = c("accession_helper", "gene_id")) %>%
  left_join(tax_merged, by = "accession_helper") %>% 
  right_join(distinct_full_2016, "qseqid") %>% 
  tidyr::separate_wider_delim(X2, delim = ";", names = c("domain", "phylum", "class","order", "family", "genus", "species")) %>%
  filter(function_diamond == "Carbon monoxide oxidation" | 
           function_diamond == "Methane oxidation" | 
           function_diamond == "hydrogenase")  %>% 
  group_by(gene, class, treatment) %>%  
  summarise(n = n()) %>% # calculate number of genes
  #filter(is.na(class)) %>% # see how many does not have taxonomy
  left_join(readxl::read_excel("gene_numbers.xlsx"), # this must be corrected by numbers from Justus!!!!
            by = "treatment") %>%
  mutate(perc = (n/gene_count)*1e6) %>% # genes per mio of genes
  filter(!is.na(class)) %>% # remove no hits from diamond tax, checked manually and even domain is NA for these genes, i.e. it might be trace gas gene but link to microbes can not be shown
  filter(gene != "FeFe") %>% # removes two genes for FeFe
  left_join(readxl::read_excel("distinct_full_taxonomy_helper.xlsx"),
            by = c("gene", "class")) %>% # join helper table to plot only most abundant
  mutate(class_helper = if_else(is.na(class_helper), true = "Others", false = class_helper)) %>%
  mutate(class_helper = str_remove(class_helper, "c__")) %>% 
  mutate(temp = fct_recode(temp, `Long−term warmed` = "warm")) %>%
  mutate(temp = fct_recode(temp, Unwarmed = "ambient")) %>%
  glimpse
```